---
title: "Analyzing Music Data"
author: "Anthony Gao"
format: html
execute:
  echo: false
---

## Setup

```{r}
#| label: setup
#| message: false
library(tidyverse)
```

## Billboard Analysis

The `billboard` dataset tracks weekly chart rankings for songs that entered the Billboard Hot 100 in the year 2000. Each row represents a single song, with columns recording the artist, track name, date entered, and rank for each week on the chart (up to 76 weeks).

### Data Preview

```{r}
#| label: billboard-preview
billboard |> 
  select(artist, track, date.entered, wk1:wk4)
```

The first four weeks of rankings show how songs debuted and moved through the chart. Most songs enter at a relatively high rank number (lower popularity) and climb over subsequent weeks.

### Distribution of Week 6 Rankings

```{r}
#| label: billboard-wk6-histogram
#| message: false
n_val <- sum(!is.na(billboard$wk6))
ggplot(billboard, aes(x = wk6)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "white") +
  labs(
    title = "Histogram of Week 6 Rankings",
    subtitle = paste0("n = ", n_val, " songs"),
    x = "Week 6 Rank",
    y = "Count"
  )
```

By week 6, many songs have already dropped off the chart entirely. Among those that remain, rankings are spread fairly evenly, with no strong clustering at any particular position.

### Data Availability Across Weeks

```{r}
#| label: billboard-missingness
billboard |> 
  select(wk1, wk4, wk10, wk20, wk40, wk76) |> 
  pivot_longer(cols = everything(), names_to = "week", values_to = "rank") |> 
  mutate(week = fct_inorder(week)) |> 
  group_by(week) |> 
  summarize(
    present = sum(!is.na(rank)),
    missing = sum(is.na(rank))
  )
```

The table reveals a steep drop-off in data availability: all 317 songs have a week-1 rank, but only a handful survive to week 40 or beyond. This reflects the reality that most songs have short chart lives.

### Chart Re-entries

```{r}
#| label: billboard-reentries
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank"
  ) |> 
  filter(!is.na(rank)) |> 
  mutate(week = as.integer(str_remove(week, "wk"))) |> 
  arrange(artist, track, week) |> 
  group_by(artist, track) |> 
  summarize(
    re_entered = any(diff(week) > 1), 
    .groups = "drop"
  ) |> 
  count(re_entered)
```

A small number of songs left the chart and later re-entered, indicating gaps in their weekly rankings. The vast majority of songs, however, had continuous (uninterrupted) chart runs.

### Song Ranking Trajectories

```{r}
#| label: billboard-trajectories
#| message: false
billboard_long <- billboard |> 
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |> 
  mutate(week = as.integer(str_remove(week, "wk")))

ggplot(billboard_long, aes(x = week, y = rank, group = interaction(artist, track))) +
  geom_line(alpha = 0.25) +
  scale_y_reverse() +
  labs(
    title = "Song Rankings Over Time",
    x = "Week",
    y = "Rank (reversed)"
  )
```

Most songs follow a common arc: they enter the chart, rise toward a peak rank within the first few weeks, then gradually fall off. A few outliers persist for 20+ weeks, showing exceptional longevity.

### Notable Songs

```{r}
#| label: billboard-notable-songs
#| message: false
# Create song-level summary
song_summary <- billboard |> 
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  ) |> 
  mutate(week = as.integer(str_remove(week, "wk"))) |> 
  group_by(artist, track) |> 
  summarize(
    first_rank = rank[which.min(week)],
    best_rank = min(rank),
    first_week_at_best = min(week[rank == best_rank]),
    total_weeks = n(),
    .groups = "drop"
  )

# Print a preview of the song-level summary
print("Preview of Song-Level Summary:")
head(song_summary)

# Identify the three notable songs
fastest_no1 <- song_summary |> 
  filter(best_rank == 1) |> 
  slice_min(first_week_at_best, n = 1)

slowest_no1 <- song_summary |> 
  filter(best_rank == 1) |> 
  slice_max(first_week_at_best, n = 1)

longest_run_top10 <- song_summary |> 
  filter(best_rank <= 10) |> 
  slice_max(total_weeks, n = 1)

# Combine and display
print("Notable Songs:")
bind_rows(
  fastest_no1 |> mutate(notable_category = "Fastest to #1"),
  slowest_no1 |> mutate(notable_category = "Slowest to #1"),
  longest_run_top10 |> mutate(notable_category = "Longest Run in Top 10")
) |> 
  select(notable_category, artist, track, first_rank, best_rank, first_week_at_best, total_weeks)
```

Three songs stand out: the fastest to reach #1, the slowest climber to eventually hit #1, and the song with the longest sustained run in the top 10. These represent different strategies for chart success — instant hits versus slow-burn growers.

### Top-10 Song Trajectories

```{r}
#| label: billboard-top10-trajectories
#| cache: true
#| message: false
# Filter for songs that reached the top 10
top10_songs <- billboard_long |> 
  group_by(artist, track) |> 
  filter(min(rank) <= 10) |> 
  ungroup() |> 
  mutate(
    highlight = case_when(
      artist == "Madonna" & track == "Music" ~ "Madonna - Music (Fastest to #1)",
      artist == "Lonestar" & track == "Amazed" ~ "Lonestar - Amazed (Slowest to #1)",
      artist == "Creed" & track == "Higher" ~ "Creed - Higher (Longest Top-10 Run)",
      TRUE ~ "Other Top-10 Songs"
    )
  )

# Plot top 10 songs, highlighting the three notable ones
ggplot() +
  geom_line(
    data = filter(top10_songs, highlight == "Other Top-10 Songs"),
    aes(x = week, y = rank, group = interaction(artist, track)),
    color = "lightgray", alpha = 0.5
  ) +
  geom_line(
    data = filter(top10_songs, highlight != "Other Top-10 Songs"),
    aes(x = week, y = rank, group = interaction(artist, track), color = highlight),
    linewidth = 1
  ) +
  scale_y_reverse() +
  scale_color_manual(
    values = c(
      "Madonna - Music (Fastest to #1)" = "#E41A1C",
      "Lonestar - Amazed (Slowest to #1)" = "#377EB8",
      "Creed - Higher (Longest Top-10 Run)" = "#4DAF4A"
    )
  ) +
  labs(
    title = "Song Trajectories for Top-10 Billboard Hits",
    subtitle = "Highlighting three notable songs",
    x = "Week",
    y = "Rank (reversed)",
    color = "Notable Songs"
  ) +
  theme_minimal()
```

The highlighted trajectories confirm the distinct patterns: Madonna's "Music" shot to #1 almost immediately, Lonestar's "Amazed" took a long, winding path to the top, and Creed's "Higher" maintained a top-10 presence for an unusually extended period.

---

## Music CSV Analysis

The `music.csv` dataset contains 10,000 songs with 35 attributes covering artist metadata (name, location, genre tags, popularity) and audio features (tempo, loudness, duration, key). Out of the 10,000 songs, only 3,742 have usable geographic coordinates, while the remaining 6,258 have placeholder coordinates (0, 0).

### Geographic Distribution by Release Year

```{r}
#| label: music-geo-year
#| message: false
#| warning: false
music <- read_csv("data/music.csv", show_col_types = FALSE)

library(maps)
world_map <- map_data("world")
music_coords <- music |> 
  filter(artist.latitude != 0 | artist.longitude != 0) |> 
  filter(song.year != 0)

ggplot() +
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group), fill = "#EAEAEA", color = "white") +
  geom_point(data = music_coords, aes(x = artist.longitude, y = artist.latitude, color = song.year), alpha = 0.6, size = 0.8) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(
    title = "Geographic Distribution of Artists by Song Release Year",
    subtitle = "Artists with known release years are concentrated in North America and Europe, with years spanning 1926 to 2010",
    x = "Longitude",
    y = "Latitude",
    color = "Year"
  )
```

After filtering out entries with missing coordinates and unknown release years (coded as 0), 1,825 songs remain. The map shows a strong concentration in the eastern United States and western Europe, with more recent releases (lighter colors) dominating. Older recordings (darker colors) appear almost exclusively in the US, reflecting the country's longer history of commercial music recording in this dataset.
